---
title: Current status and propsects of R-packages for the design of experiments
author:
  # see ?rjournal_article for more information
  - name: Emi Tanaka
    affiliation: Monash University
    address:
    - Monash University
    - Clayton campus, VIC 3800, Australia
    url: http://emitanaka.org/
    orcid: 0000-0002-1455-259X
    email:  emi.tanaka@monash.edu
abstract: >
    Re-running an experiment is generally costly and in some cases impossible due to limited resources, so the design of experiment plays a critical role in increasing the quality of experimental data.  In this article I describe the current state of the R-packages for the design of experiments through textual analysis and download trends. I discuss also the software design of widely utilised R packages in the field of experimental design and conclude with discussion of some future prospects for the field.
preamble: |
  % Any extra LaTeX you need in the preamble
  
# per R journal requirement, the bib filename should be the same as the output 
# tex file. Don't forget to rename the bib file and change this example value.
bibliography: paper.bib

output: rticles::rjournal_article
---

```{r setup, cache = FALSE, include = FALSE}
library(tidyverse)
library(rvest)
library(lubridate)
library(cranlogs)
library(glue)
library(scales)
library(colorspace)
library(tidytext)
library(pluralize)
library(kableExtra)
library(igraph)
library(ggraph)
library(ggrepel)
library(patchwork)
library(ggwordcloud)
tocache <- TRUE
knitr::opts_chunk$set(echo = FALSE, 
                      cache = FALSE,
                      cache.path = "cache/",
                      fig.align = 'center', 
                      fig.pos = 'htbp', 
                      fig.width = 6,
                      message = FALSE,
                      warning = FALSE)

theme_set(
  theme(panel.background = element_rect(fill = NA),
        panel.grid = element_line(color = "lightgray"),
        axis.text = element_text(color = "black"),
        axis.line = element_line(color = "black", size = 0.7),
        axis.ticks.length = unit(1.4, "mm"),
        axis.ticks = element_line(color = "black", size = 0.7),
        axis.title = element_text(color = "black", face = "bold"),
        strip.background = element_rect(color = "black",
                                        fill = "black"),
        strip.text = element_text(color = "white"),
        plot.title.position = "plot",
        plot.title = element_text(color = "black", hjust = 0)))
```

```{r ctv}
doe_pkgs <- ctv:::.get_pkgs_from_ctv_or_repos("ExperimentalDesign", 
                                              repos = "http://cran.rstudio.com/")[[1]]
doe_pkgs <- c(doe_pkgs, "DeclareDesign")
npkgs <- length(doe_pkgs)
```

```{r cran-data, cache = tocache}
end <- Sys.Date() - 2 # usually 1-2 days data are not available yet
start <- end - years(2) + days(2)
dldat <- cran_downloads(doe_pkgs, from = start, to = end)
```

```{r cran-data2, cache = tocache}
start2 <- end - years(5) + days(2)
dldat2 <- cran_downloads(doe_pkgs, from = start2, to = end)
```

# Introduction 

The critical role of data collection is well captured in the expression "garbage in, garbage out" -- in other words, if the collected data is rubbish then no analysis, however complex it may be, can make something out of it. Methods for data collection can be dichotomised by the type of data collected -- namely, experimental or observational -- or alternatively, categorised into experimental design or survey design.  Survey designs often, although not always, aim to collect observational data whilst experimental designs exclusively center on experimental data. This dichotomisation, to a great extent, is seen in CRAN task views where R-packages in experimental design are in \ctv{ExperimentalDesign} and R-packages in survey designs are in \ctv{OfficialStatistics}.


In the CRAN task view of \ctv{ExperimentalDesign}, there are `r npkgs` R packages for experimental design and analysis of data from experiments, henceforth referred to as "DoE packages" in this paper. The sheer quantity and variation in the output experimental design in the R-packages are arguably unmatched with any other programming languages, e.g. in Python [@python], only a handful of libraries that generate design of experiment exist (namely `pyDOE`, `pyDOE2`, `dexpy`, `experimenter` and `GPdoemd`) with limited outputs.


# Explorative Analysis

```{r download-dist, dependson = "cran-data"}
dldat_sum <- dldat %>% 
  group_by(package) %>% 
  summarise(total = sum(count)) %>% 
  arrange(desc(total)) %>% 
  mutate(rank = 1:n())

dl1plot <- ggplot(dldat_sum, aes(x = "", y = total/1000)) + 
  geom_violin() +
  geom_boxplot(width = 0.1) + 
  scale_y_log10(label = comma) + 
  labs(x = "",
       y = "Download counts in the last 2 year ('000s)", title = "(A)") + 
  geom_text_repel(data = slice_max(dldat_sum, total, n = 13), 
                  nudge_x = rep(c(0.2, -0.2, 0.2, -0.2, 0.15), c(3, 2, 2, 3, 3)),
                  size = 4,
                  aes(label = package)) + 
  theme(axis.ticks.length.x = unit(0, "mm"))


dldat_sum5 <- dldat2 %>% 
  group_by(package) %>% 
  summarise(total = sum(count)) %>% 
  arrange(desc(total)) %>% 
  mutate(rank = 1:n())

top15 <- dldat_sum %>% 
  left_join(dldat_sum5, by = "package") %>% 
  slice_max(total.x + total.y, n = 15) %>% 
  select(package, rank.x, rank.y) %>% 
  pivot_longer(-package) %>% 
  mutate(name = ifelse(name=="rank.x", "Last 2 years", "Last 5 years"))

dl2plot <- ggplot(top15, aes(name, value, group = package)) +
  geom_line() + 
  geom_point() +
  scale_y_reverse(breaks = 1:15) +
  scale_x_discrete(expand = expansion(c(0.1, 0.5))) +
  labs(x = "", y = "Ranking", title = "(B)") +
  geom_text(data = filter(top15, name=="Last 5 years"),
            aes(label = package), 
            hjust = 0, 
            nudge_x = 0.1,
            size = 4) + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

```{r dlplots, fig.cap=glue("The above graph shows the total download counts from {start} to {end} of DoE packages."), fig.width = 6}
dl1plot + dl2plot
```

```{r text, cache=tocache}
url <- "http://cran.rstudio.com/web/packages/packages.rds"
db <- readRDS(url(url)) %>% 
  as.data.frame()

doe_db <- db %>% 
  filter(Package %in% doe_pkgs) %>% 
  mutate(Description = str_replace_all(Description, "\n", " "),
         Description = str_squish(Description),
         Title = str_replace_all(Title, "\n", " "))
```

```{r fns}
singularize2 <- function(x) {
    res <- singularize(x)
    res <- ifelse(res=="bia", "bias", res)
    ifelse(res=="baye", "bayes", res)
}
stop_words_rex <- c(paste0("^", stop_words$word, "$"), "^doi", "^[0-9.]+$")
get_ngram <- function(ngram, desc = c("Title", "Description")) {
  desc <- match.arg(desc)
  words <- paste0("word", 1:ngram)
  doe_db %>% 
    unnest_tokens(word, desc, token = "ngrams", n = ngram) %>% 
    separate(word, words, sep = " ") %>% 
    mutate(across(num_range("word", 1:ngram), singularize2)) %>% 
    distinct(!!!map(c("Package", words), as.name)) %>% 
    # allow stop words in between
    filter(if_all(num_range("word", c(1, ngram)), 
                  ~!str_detect(., paste0(stop_words_rex, collapse = "|")))) %>% 
    count(!!!map(words, as.name), sort = TRUE) %>% 
    mutate(word = do.call("paste", map(words, ~eval(parse(text = .x))))) %>% 
    select(word, n) 
}

combine_dl <- function(ngram_df, desc) {
  dldat_sum %>% 
    left_join(doe_db, by = c("package" = "Package")) %>% 
    select(package, total, {{desc}}) %>% 
    mutate(bigram_title = map({{desc}}, ~filter(ngram_df, 
                                              n > 1,
                                              str_detect(tolower(.x), word)))) %>% 
    unnest_longer(bigram_title) %>% 
    unpack(bigram_title) %>% 
    group_by(word) %>% 
    summarise(n = unique(n),
              pkgs = list(package),
              total = sum(total)) %>% 
    filter(!is.na(word)) %>% 
    arrange(desc(total))
}
```

```{r ngrams, cache=tocache, dependson="fns"}
bigram_title <- get_ngram(2, "Title")
bigram_desc <- get_ngram(2, "Description")
trigram_desc <- get_ngram(3, "Description")
```

* Not much difference between title and description.
* Just go with description and combine bi & tri.

```{r wordcloud, fig.height = 3.3}
dldat_comb_bigram_title <- combine_dl(bigram_title, Title)
dldat_comb_bigram_desc <- combine_dl(bigram_desc, Title)
dldat_comb_trigram_desc <- combine_dl(trigram_desc, Title)

set.seed(1)
ggplot(rbind(dldat_comb_bigram_desc,
             dldat_comb_trigram_desc),
       aes(label = word, size = total, color = n)) +
  geom_text_wordcloud(shape = "square",
                      show.legend = TRUE) +
  scale_size_area(max_size = 8) +
  guides(size = FALSE) +
  scale_color_continuous(breaks = 1:11) +
  theme(legend.position = "bottom")
```




## What are the common types of experimental designs?

## How do packages interplay with each other?

## Which packages are widely utilised?

# Software Design

# Propsects and Discussion

\newpage



```{r cran-titles}
url <- "http://cran.rstudio.com/web/packages/packages.rds"
db <- readRDS(url(url)) %>% 
  as.data.frame()
```

```{r bigram}
stop_words_ext <- c(stop_words$word, "doi")

doe_db <- db %>% 
  filter(Package %in% doe_pkgs) %>% 
  mutate(Description = str_replace_all(Description, "\n", " "),
         Description = str_squish(Description),
         Title = str_replace_all(Title, "\n", " "))

bigram_tab <- function(data, col) {
  data %>% 
    unnest_tokens(word, {{col}}, token = "ngrams", n = 2) %>% 
    separate(word, c("word1", "word2"), sep = " ") %>% 
    mutate(word1 = singularize(word1),
           word2 = singularize(word2)) %>% 
    # don't count the same bigram within the same package
    distinct(Package, word1, word2) %>% 
    filter(!word1 %in% stop_words_ext,
           !word2 %in% stop_words_ext,
           !str_detect(word1, "^[0-9.]+$"),
           !str_detect(word2, "^[0-9.]+$")) %>% 
    count(word1, word2, sort = TRUE)  
}
```

```{r bigram-title, results="asis"}
btitle <- bigram_tab(doe_db, Title) %>% 
  filter(n > 3) %>% 
  mutate(word = paste(word1, word2)) %>% 
  select(word, n) %>% 
  kbl(booktabs = TRUE, linesep = "")

bdesc <- bigram_tab(doe_db, Description) %>% 
  filter(n > 4) %>% 
  mutate(word = paste(word1, word2)) %>% 
  select(word, n) %>% 
  kbl(booktabs = TRUE, linesep = "")

cat(c("\\begin{table}[h] \\centering ", 
      btitle,
    "\\hspace{1cm} \\centering ",
      bdesc,
    "\\caption{My tables} \\end{table}")) 
```

```{r}
doe_imports <- doe_db %>% 
  mutate(Depends = str_replace_all(Depends, "\n", " "),
         Depends = str_replace_all(Depends, fixed("("), " ("),
         Imports = str_replace_all(Imports, "\n", " "),
         Imports = str_replace_all(Imports, fixed("("), " ("),
         imports = str_c(Depends, Imports, sep = ","),
         imports = str_split(imports, ","),
         imports = map(imports, ~{
           str_squish(.x) %>% 
             word() %>% 
             .[.!=""]}
         ),
         imports_doe = map(imports, ~.x[.x %in% doe_pkgs])) %>% 
  select(Package, imports_doe) %>% 
  unnest_longer(imports_doe) %>% 
  filter(!is.na(imports_doe)) %>% 
  rename(from = imports_doe, to = Package) %>% 
  select(from, to)
```

```{r doe-network, eval = FALSE}
graph_from_data_frame(doe_imports) %>% 
  ggraph(layout = 'fr') +
  geom_edge_link(aes(start_cap = label_rect(node1.name),
                     end_cap = label_rect(node2.name)), 
                 arrow = arrow(length = unit(2, 'mm')),
                 color = "#79003e") + 
  geom_node_text(aes(label = name),
                 color = "#79003e") +
  theme(panel.background = element_rect(fill = "#f6e5ee",
                                        color = "#79003e"),
        plot.margin = margin(20, 20, 20, 20))
```



\CRANpkg{agricolae}



## Helping info to get started

Introductory section which may include references in parentheses
[@R], or cite a reference such as @R in the text.

## Section title in sentence case

Let's check fi this works Figure \@ref(fig:Rlogo).

This section may contain a figure such as Figure \ref{fig:Rlogo}.




## Summary

This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).

### About this format and the R Journal requirements

`rticles::rjournal_article` will help you build the correct files requirements: 

* A R file will be generated automatically using `knitr::purl` - see
https://bookdown.org/yihui/rmarkdown-cookbook/purl.html for more information.
* A tex file will be generated from this Rmd file and correctly included in
`RJwapper.tex` as expected to build `RJwrapper.pdf`.
* All figure files will be kept in the default rmarkdown `*_files` folder. This
happens because `keep_tex = TRUE` by default in `rticles::rjournal_article`
* Only the bib filename is to modifed. An example bib file is included in the
template (`RJreferences.bib`) and you will have to name your bib file as the
tex, R, and pdf files.
