---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(tidyverse)
paper <- here::here("paper.Rmd")
meta <- rmarkdown::yaml_front_matter(paper)
```


# `r meta$title`

This is the repo for the paper titled "`r meta$title`". You can find the arxiv version of the paper in `paper/arxiv/arxiv.pdf`. To reproduce the results of the paper, you will need to run the analysis by calling 

```r
targets::tar_make()
```

The paper uses the developmental version of the [`cranscrub` package in the branch `paper-doe-review`](https://github.com/numbats/cranscrub/tree/paper-doe-review) which can be installed by 

```r
remotes::install_github("numbats/cranscrub", ref = "paper-doe-review")
```


## Authors

```{r, echo = FALSE}
map_chr(meta$author, ~{
  paste0("- ", .x$name, ", ", .x$affiliation)
}) %>% 
  cli::cli_li() 
```

## Abstract

`r meta$abstract`


## Supplementary material

The supplementarial material can be found [here](https://emitanaka.org/paper-DoE-review/supp.pdf).